Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-04-13 manuscript/manuscript.tex

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7 
trajin md-01.nc 1 1 1 
autoimage !(:WAT) origin 
strip :WAT 
trajout biotin-centered.pdb 
run
\end{lstlisting}
In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fifth line. 
Finally, we start the analysis with the "run" command.

@@

Here, we do not include any sampling of the conformational states.
It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}[style=cpptraj]
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
Adapt the above input file for biotin by stripping the molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.

@@

\begin{table}[h]
	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$. ``Diff'' is calculated as 
	$E^\text{internal} + \Delta E^\text{GIST} - T\Delta S^\text{scaled}$. ``total'' is calculated as 
	``complex''-``streptavidin'' - ``biotin''.}\label{tab_dg_monomers_dimer}
	\small
	\begin{tabular}{@{}lrrrrr@{}}
		\toprule
		System       & $E^{internal}$ & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
		\midrule
		complex      & -1303.4 & -361.2 & -233.1 & -139.9 & -1524.7 \\
		streptavidin & -1180.7 & -360.8 & -242.4 & -145.4 & -1396.1 \\
		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 &  -101.0 \\
		\midrule
		total         & -98.7   &   97.7 &   44.4 &   26.6 &   -27.6 \\
		\bottomrule
	\end{tabular}
\end{table}
Even though streptavidin-biotin is known to be a very stable complex, the free solvation energy ($\Delta E^{GIST}-T\Delta S^{scaled}$) favors the dissociation.
This is expected: the binding of biotin to streptavidin is facilitated over a large number of H-bonds and over polar interactions. 
As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
In this case, we find that the changes in the energy of solvation and of the internal energy $\Delta E^{internal}$ more or less cancel out, indicating that binding is largely entropy-driven in this case.
At first, this is surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions \cite{mpye2020-biotin-itc,hyre2006-biotin-itc}.
However, we do not take the conformational transition of the streptavidin binding pocket into account.

@@

\begin{lstlisting}[style=pymol]
load output/streptavidin/gist.pdb, streptavidin
as surface
color gray70, streptavidin
load output/streptavidin/gist-gO.dx, gO
isomesh gO_high, gO, 2
color slate, gO_high
\end{lstlisting}
A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
molecule, is found in Figure~\ref{fig-binding-pocket}.
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density (i.e., bulk)}
	\label{fig-streptavidin_gO}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
	\caption{GIST solvent energy (based on $E_{sw} + 2E_{ww}$).}
	\label{fig-binding-pocket-energy}
    \end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
		\caption{GIST solvent entropy}
		\label{fig-binding-pocket-entropy}
	\end{subfigure}
	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\cubic\angstrom}, except for 
	the positive energy, where they are divided by two to visually account for its overall lower density. The negative solvent energy is shown in green, positive solvent energy in red and negative entropy in blue. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
	\label{fig-binding-pocket}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/sasa-simple.png}
	\caption{SAS for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
\end{figure}
\subsection{Defining the binding pocket}
Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
This is significant especially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
For example, voxels that are located within the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
In section~\ref{sec:binding_contributions}, we show how to integrate over regions such as the binding pocket. To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0. Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.

>>>>

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While all MD simulations require a large computational cost, GIST is more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis highlights details and considerations relevant to more advanced GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7 
trajin md-01.nc 1 1 1 
autoimage !(:WAT) origin 
strip :WAT 
trajout biotin-centered.pdb 
run
\end{lstlisting}
In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
Finally, we start the analysis with the "run" command.

@@

Here, we do not include any sampling of the conformational states.
It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}[style=cpptraj]
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.

@@

\begin{table}[h]
	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$. ``Diff'' is calculated as 
	$E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$. ``total'' is calculated as 
	``complex''-``streptavidin'' - ``biotin''.}\label{tab_dg_monomers_dimer}
	\small
	\begin{tabular}{@{}lrrrrr@{}}
		\toprule
		System       & $E^{internal}$ & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
		\midrule
		complex      & -1303.4 & -361.2 & -233.1 & -139.9 & -1524.7 \\
		streptavidin & -1180.7 & -360.8 & -242.4 & -145.4 & -1396.1 \\
		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 &  -101.0 \\
		\midrule
		total         & -98.7   &   97.7 &   44.4 &   26.6 &   -27.6 \\
		\bottomrule
	\end{tabular}
\end{table}
Even though streptavidin-biotin is known to be a very stable complex, the free solvation energy ($\Delta E^{GIST}-T\Delta S^{scaled}$) favors the dissociation.
This is expected: the binding of biotin to streptavidin is facilitated over a large number of H-bonds and over polar interactions. 
As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
In this case, we find that the changes in the energy of solvation and of the internal energy $\Delta E^{internal}$ more or less cancel out, indicating that binding is largely entropy-driven in this case.
At first, this is surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions \cite{mpye2020-biotin-itc,hyre2006-biotin-itc}.
However, we do not take the conformational transition of the streptavidin binding pocket into account.

@@

\begin{lstlisting}[style=pymol]
load output/streptavidin/gist.pdb, streptavidin
as surface
color gray70, streptavidin
load output/streptavidin/gist-gO.dx, gO
isomesh gO_high, gO, 2
color slate, gO_high
\end{lstlisting}
A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
molecule, is found in Figure~\ref{fig-binding-pocket}.
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density (i.e., bulk)}
	\label{fig-streptavidin_gO}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
	\caption{GIST solvent energy (based on $E_{sw} + 2E_{ww}$).}
	\label{fig-binding-pocket-energy}
    \end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
		\caption{GIST solvent entropy}
		\label{fig-binding-pocket-entropy}
	\end{subfigure}
	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\cubic\angstrom}, except for 
	the positive energy, where they are divided by two to visually account for its overall lower density. The negative solvent energgy is shown in green, positive solvent energy in red and negative entropy in blue. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
	\label{fig-binding-pocket}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/sasa-simple.png}
	\caption{SAS for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
\end{figure}
\subsection{Defining the binding pocket}
Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
For example, voxels that are located within the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
In section~\ref{sec:binding_contributions}, we show how to integrate over regions such as the binding pocket. To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0. Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.
